Non-Vertical Cultural Transmission and the\\ Evolution of Cooperation

### 2020-08-19 ms/main.tex
revised title, figure captions, corollay 5, names of variables and more

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex
\graphicspath{ {./Figures/} }
\usepackage{subcaption}
\usepackage{float}

@@

@@

% Title page
\title{Cultural Transmission Can Explain the Evolution of Cooperation}

@@

\author[1]{Dor Cohen}
\author[2]{Ohad Lewin-Epstein}
\author[3]{Marcus W. Feldman}
\author[a,*]{Yoav Ram}

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya, Israel}
\affil[2]{School of Plant Sciences and Food Security, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Biology, Stanford University, Stanford, CA}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

\emph{Kin selection} theory posits that natural selection is more likely to favor cooperation between more closely related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}, who showed that an allele that determines cooperative behavior will increase in frequency if the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the relatedness, $r$, between the recipient and the actor. This relatedness coefficient $r$ measures a probability that an allele sampled from the cooperator is identical by descent to one at the same locus in the recipient.
This condition is  known as ``Hamilton's rule'':
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

\citet{Eshel1982}  studied a related model for the evolution of cooperative behavior under vertical transmission. Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
} 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:assortative_meeting}
c < b \cdot m\,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation. 
Here $m$ in (\ref{eq:assortative_meeting}) takes the role of the relatedness $r$  in (\ref{eq:hamilton_rule}).

%These theories assume that cooperation is genetically determined, which raises the question: \emph{Is it possible that cooperation is determined by non-genetic factors?}
%Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.

Here we study the evolution of a cooperative behavior that is subject to \emph{cultural transmission},
 which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission and
showed that under vertical (parent-to-offspring) cultural transmission, the conditions for the initial increase of altruism entails a modification of Hamilton's rule in the cases  of parent-to-offspring or sib-to-sib altruism.

@@

We hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation, and 
to test this hypothesis we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is an advantageous behavior and will be cooperative in the future. 
Our cultural evolution models  include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not, and that treatment of cooperation as a cultural, rather than a genetic, trait can lead to a better understanding of its evolutionary dynamics.

@@

Consider a very large population whose members are characterized by their phenotype $\phi$, which can be of two types, $\phi=A$ for cooperators or $\phi=B$ for defectors.
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$. 
Following~\citet{ram2018evolution}, given that the parent phenotype is $\phi$ and assuming uni-parental inheritance, % yr: cite Zefferman Behav Ecol 2016
the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is

@@

Individuals are assumed to interact according to a \emph{prisoner's dilemma}.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $c<b$. \textbf{\autoref{table:prisoner_payoff}} shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

\bigskip
\begin{center}[TABLE 1 HERE]\end{center}
%%% Table: payoff matrix

\bigskip
Social interactions occur randomly:
two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$.

Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
The assortment parameter $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
Horizontal transmission is not always successful, as one partner may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\textbf{\autoref{table:interactions}}).

\bigskip
\begin{center}[TABLE 2 HERE]\end{center}
%%% Table: payoff matrix

\bigskip

@@

The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \textbf{\autoref{table:prisoner_payoff}} and \textbf{\autoref{table:interactions}}, and the population mean fitness is
\begin{equation} \label{eq:mean_fitness}
\bar{w} =  1 + \hat{p}(b-c).
\end{equation}

@@

%%% Table: interactions

%%% Figure: model illustration
\begin{figure}[thb]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}
%%%

@@

Let the initial frequency of the alternative phenotype be $\tilde{p}_0$ and denote the cost boundaries by
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\gamma_1 = \frac{b \alpha T_A + (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}
\end{equation}
Then, applying \ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2}, we  summarize the possible outcomes which are illustrated in \autoref{fig:results2}:
%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item Fixation of cooperation,
	if $T_A>T_B$ and $c<\gamma_1$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0>\tilde{p}^*$; or
	if $T_A<T_B$ and $c<\gamma_2$.
\item Fixation of defection,
	if $T_A>T_B$ and $\gamma_2<c$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0<\tilde{p}^*$; or
	if $T_A<T_B$ and $\gamma_1<c$.
\item Coexistence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and $\gamma_2<c<\gamma_1$.
\end{enumerate}

@@

%%% Figure: Result 2 Boundaries

\begin{figure}[htb]
  \centering
    \includegraphics[scale=0.75]{figure2a.pdf}
    \includegraphics[scale=0.7]{figure2b.pdf}
    \includegraphics[scale=0.7]{boundiries_withAlpha_y_axis.pdf}
    \caption{\textbf{Vertical and horizontal transmission.} The figure illustrates (\textbf{a-b}): combinations of horizontal bias, $T_A-T_B$, and cost of cooperation, $c$; (\textbf{c}) combinations of horizontal bias, $T_A-T_B$, and assortative meeting, $\alpha$, 
    that lead to either global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial frequencies (blue), or coexistence of cooperation and defection (yellow). 
    In (\textbf{a-b}): the blue and orange curves show the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}).
  Here, benefit of cooperation is $b=1.3$, horizontal transmission of cooperation $T_A=0.4$, (\textbf{a-b}): $c$ and $T_B$ vary on the y- and x-axes; (\textbf{c}): $\alpha$ and $T_B$ vary on the y- and x-axes.
   (\textbf{a}): With $\alpha = 0.7 > 0$, coexistence is possible (yellow). (\textbf{b}): With $\alpha=0$, coexistence is not possible. (\textbf{c}): with $c = 0.35$ coexistence is possible.}
    \label{fig:results2}
\end{figure}
%%%

%%% Figure: coexistence map
\begin{figure}[htb]
\centering
\includegraphics[scale=1]{coexistence.pdf}
\caption{\textbf{Stable coexistence between cooperation and defection.} The frequency of the cooperative phenotype $A$ among juveniles in the next generation $\hat{p}'$ is plotted in blue as a function of the frequency in the current generation $\hat{p}$ for $T_A<T_B$ and $\gamma_2<c<\gamma_1$. The line of $\hat{p}'=\hat{p}$ is in dashed black. The curve and dashed line intersect at the equilibrium $\hat{p}^*$. 
The blue curve is above the black dashed line$\hat{p}<\hat{p}^*$, so that the frequency increases towards $\hat{p}^*$. The blue curve is below the black dashed line when $\hat{p}>\hat{p}^*$, so that the frequency decreases towards $\hat{p}^*$.
}
\label{fig:coexistence}
\end{figure}

@@

\begin{corollary}[No assortment of transmission and cooperation]
If $\alpha=0$ and there is horizontal bias for cooperation ($T_A>T_B$) and
\textbf{(1)} the cost is low compared to the bias ($c<(T_A-T_B)/(1-T_B)$), then cooperation will fix from any positive frequency; or
\textbf{(2)} the cost is low compared to the benefit ($c<(1+b)(T_A-T_B)(1-T_B)$), then cooperation will fix if the initial frequency is high enough ($\tilde{p}_0 > \tilde{p}^*$).
\end{corollary}

@@

\begin{corollary}[Complete assortment of transmission and cooperation]
When $\alpha=1$, there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found from inequality \ref{eq:vert_hori_alpha1_condition_proof}, namely
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
With complete assortment, in inequality \ref{eq:vert_hori_alpha1_condition_proof}
 horizontal transmission  occurs together with the cooperative interaction. The same occurs in \citet{lewin2017microbes}, and therefore this corollary is equivalent to their result (see their eq.~1).

In terms of the cost boundaries, inequality \ref{eq:vert_hori_alpha1} is equivalent to $c<\gamma_1$, and if $T_A>T_B$ then that suffices for fixation of cooperation. If $T_B>T_A$ then $\gamma_2(\alpha=1)<0$ and again, (\ref{eq:vert_hori_alpha1}) is sufficient for increase in the frequency of $A$.
Inequality \ref{eq:vert_hori_alpha1} can be written as
\begin{equation} \label{eq:vert_hori_alpha1_effective}
1 - (1-c)(1-T_B) < (1+b) T_A ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
In the interaction between a cooperator and a defector.
$(1-c)(1-T_B)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-c)(1-T_B)$ is the probability that either the cooperator becomes a defector, or that it fails to reproduce.
This is the effective cost for cooperation from this interaction, while
$(1+b) T_A$ is the probability that the defector becomes cooperative and reproduces, which
 is the effective benefit for cooperation from this interaction.
Thus inequality (\ref{eq:vert_hori_alpha1}) entails that cooperation can evolve if the effective cost for cooperation is less than the effective benefit.

\begin{figure}[htb]
  \centering
    \includegraphics[scale=0.75]{figure3a.pdf}
%    \caption{$v=1$, $T_A=T_B=T$, $\alpha \neq 0$}
  %  \label{fig:results_a}
    \includegraphics[scale=0.75]{figure3b.pdf}
%    \caption{$v=1$, $\alpha = 0$}
 %   \label{fig:results_b}
    \includegraphics[scale=0.75]{figure3c.pdf}
%    \caption{$v=0$}
%  \label{fig:results}
  \caption{
  \textbf{Numerical results for cultural evolution of cooperation.}
  Shown are dynamics of \textbf{(a-b)} $\tilde{p}$, the frequency of parents with cooperative phenotype $A$; \textbf{(c)} $p$, the frequency of adults with cooperative phenotype $A$.
  The figure demonstrates fixation of cooperation (green), extinction of cooperation (blue), and stable coexistence of cooperators and defectors (orange).
  Parameters: (a) $v=1$, $T_A=T_B=T$, $\alpha \neq 0$; (b) $v=1$, $\alpha = 0$; (c) $v=0$.
  % TODO add all parameters.
  % TODO remove legends, increase fonts, explain in caption instead of legend.
  }
  \label{fig:results}
\end{figure}

@@

%There is an ongoing debate about the extent to which kin selection explains the evolution of cooperation and altruism.
%For example, it has been suggested that it can explain the cooperative behavior of worker castes of eusocial insects like the honey bee. % TODO REF
%The most significant argument against kin selection is that in some cases cooperation among unrelated individuals appears to have evolved~\citep{wilson2005kin}.
%Therefore, other theories have been develop to explain the evolution of cooperation and altruism.

>>>>

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex
\graphicspath{ {./Figures/} }
\usepackage[labelfont=bf]{caption}
\usepackage{float}

@@

% autoref
\def\equationautorefname{Eq.}

@@

% Title page
\title{Non-Vertical Cultural Transmission and the\\ Evolution of Cooperation}

@@

\author[1]{Dor Cohen}
\author[2]{Ohad Lewin-Epstein}
\author[3]{Marcus W. Feldman}
\author[1,4,*]{Yoav Ram}

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya, Israel}
\affil[2]{School of Plant Sciences and Food Security, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Biology, Stanford University, Stanford, CA}
\affil[4]{School of Zoology, Tel Aviv University, Tel Aviv, Israel}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

\emph{Kin selection theory} posits that natural selection is more likely to favor cooperation between more closely related individuals.
The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}, who showed that an allele that determines cooperative behavior will increase in frequency if the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the relatedness, $r$, between the recipient and the actor. This relatedness coefficient $r$ measures a probability that an allele sampled from the cooperator is identical by descent to one at the same locus in the recipient.
This condition is  known as \emph{Hamilton's rule}:
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

\citet{Eshel1982} studied a related model for the evolution of cooperative behavior under vertical transmission. Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
} 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:eshel1982}
c < b \cdot m \,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation. 
Here $m$ in \autoref{eq:eshel1982} takes the role of the relatedness $r$  in \autoref{eq:hamilton_rule}.

Here we study the evolution of a cooperative behavior that is subject to \emph{cultural transmission}, which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission.
They showed that under a combination of genetic and vertical (parent-to-offspring) cultural transmission, the conditions for the initial increase of altruism entails a modification of Hamilton's rule in the cases of parent-to-offspring or sib-to-sib altruism.
For example, if the fidelity of cultural transmission of altruism is $\varphi$, then the condition for evolution of altruism in the case of sib-to-sib altruism is \citep[Eq.~16]{feldman1985gene}
\begin{equation} \label{eq:feldman1985}
c < b \cdot \varphi - \frac{1-\varphi}{\varphi} \,.
\end{equation}

@@

% TODO
We hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation, and 
to test this hypothesis we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is an advantageous behavior and will be cooperative in the future. 
Our cultural evolution models  include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not, and that treatment of cooperation as a cultural, rather than a genetic, trait can lead to a better understanding of its evolutionary dynamics.

@@

Consider a very large population whose members are characterized by their phenotype $\phi$, which can be of two types, $\phi=A$ for cooperators or $\phi=B$ for defectors.
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$. 
Following~\citet{ram2018evolution}, given that the parent phenotype is $\phi$ and assuming uni-parental inheritance, % TODO cite Zefferman Behav Ecol 2016
the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is

@@

Individuals are assumed to interact according to a \emph{prisoner's dilemma}.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $c<b$. \autoref{table:prisoner_payoff} shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

%%% Table: payoff matrix
\bigskip
\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\caption{\textbf{Payoff matrix for prisoner's dilemma.}
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $b>c>0$.
}
\label{table:prisoner_payoff}
\end{table}
\bigskip
%%%

Social interactions occur randomly:
two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$.

Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\autoref{fig:horizontal}).
The assortment parameter $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
Horizontal transmission is not always successful, as one partner may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}).

%%% Table: interactions
\bigskip
\begin{table}[]
\begin{tabular}{@{}llllll@{}}
\toprule
\multirow{2}{*}{Phenotype $\phi_1$} &
  \multirow{2}{*}{Phenotype $\phi_2$} &
  \multirow{2}{*}{Frequency} &
  \multirow{2}{*}{Fitness of $\phi_1$} &
  \multicolumn{2}{l}{$P(\phi_1=A)$ via horizontal transmission:} \\ \cmidrule(l){5-6} 
    &     &                      &         & from partner, $\alpha$ & from population, $(1-\alpha)$ \\ \cmidrule(r){1-6}
$A$ & $A$ & $\hat{p}^2$          & $1+b-c$ & 1                      & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$A$ & $B$ & $\hat{p}(1-\hat{p})$ & $1-c$   & $1-T_B$                & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$B$ & $A$ & $\hat{p}(1-\hat{p})$ & $1+b$   & $T_A$                  & $\hat{p} T_A$                 \\
$B$ & $B$ & $(1-\hat{p})^2$      & $1$     & $0$                    & $\hat{p} T_A$                 \\ \bottomrule
\end{tabular}
\caption{\textbf{Interaction frequency, fitness, and transmission probabilities.}}
\label{table:interactions}
\end{table}
\bigskip
%%%

@@

The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{table:prisoner_payoff} and \autoref{table:interactions}, and the population mean fitness is
\begin{equation} \label{eq:mean_fitness}
\bar{w} =  1 + \hat{p}(b-c).
\end{equation}

@@

@@

Let the initial frequency of the alternative phenotype be $\tilde{p}_0$ and denote the cost boundaries by
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\gamma_1 = \frac{b \alpha T_A + (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}
\end{equation}
Then, applying Eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2}, we  summarize the possible outcomes which are illustrated in \autoref{fig:results2}:
%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item Fixation of cooperation,
	if $T_A>T_B$ and $c<\gamma_1$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0>\tilde{p}^*$; or
	if $T_A<T_B$ and $c<\gamma_2$.
\item Fixation of defection,
	if $T_A>T_B$ and $\gamma_2<c$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0<\tilde{p}^*$; or
	if $T_A<T_B$ and $\gamma_1<c$.
\item Coexistence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and $\gamma_2<c<\gamma_1$.
\end{enumerate}

@@

@@

\begin{corollary}[No assortment of transmission and cooperation]
If $\alpha=0$ and there is horizontal bias for cooperation ($T_A>T_B$) and
(1) the cost is low compared to the bias ($c<(T_A-T_B)/(1-T_B)$), then cooperation will fix from any positive frequency; or
(2) the cost is low compared to the benefit ($c<(1+b)(T_A-T_B)(1-T_B)$), then cooperation will fix if the initial frequency is high enough ($\tilde{p}_0 > \tilde{p}^*$).
\end{corollary}

@@

\begin{corollary}[Perfect assortment of transmission and cooperation]
When $\alpha=1$, there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found from inequality \ref{eq:vert_hori_alpha1_condition_proof}, namely
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
With perfect assortment, in inequality \ref{eq:vert_hori_alpha1_condition_proof}
 horizontal transmission  occurs together with the cooperative interaction. The same occurs in \citet{lewin2017microbes}, and therefore this corollary is equivalent to their result (see their eq.~1).

In terms of the cost boundaries, inequality~\ref{eq:vert_hori_alpha1} is equivalent to $c<\gamma_1$, and if $T_A>T_B$ then that suffices for fixation of cooperation. If $T_B>T_A$ then $\gamma_2(\alpha=1)<0$ and again, inequality~\ref{eq:vert_hori_alpha1} is sufficient for increase in the frequency of $A$.
Inequality~\ref{eq:vert_hori_alpha1} can be written as
\begin{equation} \label{eq:vert_hori_alpha1_effective}
1 - (1-c)(1-T_B) < (1+b) T_A ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
In the interaction between a cooperator and a defector $(1-c)(1-T_B)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-c)(1-T_B)$ is the probability that either the cooperator becomes a defector, or that it fails to reproduce.
This is the effective cost for cooperation from this interaction, while $(1+b) T_A$ is the probability that the defector becomes cooperative and reproduces, which is the effective benefit for cooperation from this interaction.
Thus inequality~\ref{eq:vert_hori_alpha1} entails that cooperation can evolve if the effective cost for cooperation is less than the effective benefit.

Given the previous two corollaries for no assortment and perfect assortment ($\alpha=0$ and $\alpha=1$), it is interesting to examine the general effect of assortment on the evolution of cooperation.
We denote the assortment boundaries by
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting}
  a_1 = \frac{c(1-T_B)-(1+b)(T_A-T_B)}{b\cdot T_B}, \quad
  a_2 = \frac{c(1-T_B)-(T_A-T_B)}{b\cdot T_A}.
\end{aligned}\end{equation}
Applying Eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2}, we have the following corollary.
\\

\begin{corollary}[Intermediate assortment of transmission and cooperation]
The cooperative phenotype will increase from rarity if the assortment is high enough, or specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

\autoref{fig:results2}c demonstrates these conditions. With horizontal bias for cooperation ($T_A>T_B$; positive x-axis) cooperation fixes in the population from any initial positive frequency (green) if $\alpha$ (y-axis) is above the orange line ($a_1$), or if initially present at a high enough frequency (blue) if $\alpha$ is above the blue line ($a_2$).
With horizontal bias for defection ($T_B>T_A$; negative x-axis) cooperation fixes if $alpha$ is above the blue line ($a_2$), but can be maintained in coexistence with defection (yellow) if $\alpha$ is above the orange line ($a_1$). Without horizontal bias ($T_A=T_B$) fixation occurs if $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\autoref{}).

@@

\citet{Eshel1982} showed that with assortative meeting, namely, a probability $m$ that individuals interact  within their phenotypic group, cooperation can evolve if $c < b \cdot m$.
Our results highlight another possibility for assortatment, namely, individuals interacting at rate $\alpha$ with their cultural partners, resulting in horizontal transmission.
We show that high levels of assortment  significantly increase the potential for evolution of cooperation. With a high enough $\alpha$, cooperation can increase when initially rare (although it will not fix) even when there is horizontal bias against cooperation ($\alpha > \big(c(1-T_B) + (T_B-T_A)\big) / b T_A$, see Result~\autoref{result:vert_hori})

\citet{feldman1985gene} studied the dynamics of an altruistic phenotype with vertical cultural transmission and a gene that modifies the transmission of the phenotype. Their results are very sensitive to this genetic modification: without it, the conditions for invasion of the altruistic phenotype reduce to Hamilton's rule.
Further work is needed to include such genetic modification of cultural transmission to our model.
